---
  title: "2021-12-08 Img collages"
  author: "Sari Giering"
output: html_document
editor_options: 
  chunk_output_type: console
---
  
  ```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Background
I'm exploring whether I can make a fun collage for the press release.

I got the code from rviews.rstudio.com/2020/02/13/photo-mosaics-in-r/.

The post actually only explains how to do it, and the actual code is in the 
package description.

First you have to convert your images into tiles.
There was an issue with this function and I had to modify it (after line 15) using
the fix command. The actual code is provided in the comment section
of the blog post (screen shot).
The typed version is here in the markdown file.

Also, the code only works on jpeg files, so the initial step was to
convert the images from tif to jpeg. I only used the first 5000 images
and hope it'll be fine.

```{r load-images}
  install.packages("RsimMosaic")
  library(RsimMosaic)
  library(tiff)
  library(jpeg)

  # convert some tifs to jpeg
  for(i in 1:5000){
    IMG  <- index_all$filename[i]
    OUT  <- sub(".tif", ".jpeg", IMG) 
    img  <- readTIFF(paste("/scratch/data/images/", IMG, sep = ""))
    writeJPEG(img, target = paste("/home/sari.giering/Documents/images/", OUT, sep = ""), quality = 1)
  }

  # path for images
  path_to_img<- file.path("/home/sari.giering/Documents/images")
  
  # path to tiles
  path_to_tiles<- file.path("/home/sari.giering/Documents/tiles")

  createTiles2 <- createTiles
  createTiles3 <- createTiles2
  fix(createTiles3)
  
  # add following after line 15
  # print(filenameArray[i])
  
  # add this after previous line
    #      tryCatch(
    #         expr = { # Your code goes here...
    #             img <- jpeg::readJPEG(paste(inPath, "/", filenameArray[i], 
    #                                                 sep = ""))
    #             intrpArray <- array(dim = c(tileHeight, tileHeight/dim(img)[1] * 
    #                                             dim(img)[2], 3))
    #             intrpArray[, , 1] <- bilinearInterpolator(img[, , 1], 
    #                                                       dim(intrpArray)[1], dim(intrpArray)[2])
    #             intrpArray[, , 2] <- bilinearInterpolator(img[, , 2], 
    #                                                       dim(intrpArray)[1], dim(intrpArray)[2])
    #             intrpArray[, , 3] <- bilinearInterpolator(img[, , 3], 
    #                                                       dim(intrpArray)[1], dim(intrpArray)[2])
    #             jpeg::writeJPEG(intrpArray[1:tileHeight, 1:tileHeight, 
    #             ], paste(outPath, filenameArray[i], sep = ""))
    #         },
    #         error = function(e){
    #             # (Optional)
    #             # Do this if an error is caught...
    #         },
    #         warning = function(w){
    #             # (Optional)
    #             # Do this if an error is caught...
    #         },
    #         finally = {
    #             # (optional)
    #             # Do this at the end before quitting the tryCatch structure...
    #         }
    #     )
    # }
  
  # create tiles
  createTiles3(path_to_img, path_to_tiles, tileHeight = 30)
```

```{r make_mosaic}
  # set file name for base image
  setwd("/home/sari.giering/Downloads")
  path_to_origin <- file.path("/home/sari.giering/Downloads", "copepod_50.jpg")
  
  # load images for tiles
  path_to_tiles <- file.path("/home/sari.giering/Documents/tiles")
  
  # path for output
  path_to_output <- file.path("/home/sari.giering/Downloads/mosaic_4.jpg")
  
  composeMosaicFromImageRandom(path_to_origin, path_to_output, path_to_tiles, removeTiles = FALSE)
```