---
title: "2021-12-08 Img collages"
author: "Sari Giering"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Background
It would be great to see a few example images from each label3 class to get a feeling of class diversity. The idea is to make an image collage for each class comprised of the four largest and smallest images in each class.

# Access image file info
I have a list of the image info and the taxonomic labels. list I want to add the file size of each image to this list. To retrieve file size info, I will make a list of the files in the data folder.

Note, there are some duplicates (from an earlier augmentation effort) in the folder. I will remove those for this analysis.

  - index_all: all images
  - index: unique images (no duplicates)

```{r read_file_info}
  # read index file
  index_all <- read.csv("/scratch/data/index.csv", header = TRUE)
  
  # retrieve file info
  objs <- file.info(list.files("/scratch/data/images/", full.names = TRUE))
  knitr::kable(head(objs))
  
  # clean file names
  objs$filename_long  <- rownames(objs)
  objs$filename       <- sub("/scratch/data/images//", "", objs$filename_long)
  
  # add image size to index file
  index_all$size <- objs$size[match(index_all$filename, objs$filename)]
```

So, I noticed that some images have duplicate names distinguished by unique suffixes. After inspection, it appears that 574 images have been already augmented by simple rotation. I will exclude these.

```{r top-bot-5}
  # plot image
  library(bioimagetools)

  # plot example of duplicate images
  par(pty = "s", mfrow = c(2,2), mar = c(0,0,0,0))
   for(i in grep("Pia1\\.2017\\-10\\-03\\.1726\\+N00359755", index_all$filename)) {
    IMG_NAME <- paste("/scratch/data/images/", index_all$filename[i], sep="")
    img      <- readTIF(IMG_NAME)
  
    img(img, z = 1, col = "rgb")
   }
  par(pty = "m")

  # number of rotated images
  table(grepl("fxy\\.tif", index_all$filename))
  table(grepl("fx\\.tif",  index_all$filename))
  table(grepl("fy\\.tif",  index_all$filename))
  
  # exclude duplicate images
  index <- index_all[!(grepl("fxy\\.tif", index_all$filename)| 
                       grepl("fx\\.tif", index_all$filename) |
                       grepl("fy\\.tif", index_all$filename) ), ]
```

Extract largest and smallest 5 images of each group.

I then make a collage for each type and save these. To each collage, I add a scalebar, which I made in Adobe Illustrator. I also add the name of the label3 class and the number of images within the class.

Note, there are 39 groups in label3.

I commented out the image saving command here. Please enable if you want to regenerate the images.

```{r top5}
  library(dplyr)

  # sort all images according to size, group by label3 and only take top 5 each.
  largest_5 <- index %>%
    arrange(desc(size)) %>%
    group_by(label3) %>% slice(1:5)
  
  smallest_5 <- index %>%
    arrange(size) %>%
    group_by(label3) %>% slice(1:5)
    
  # extract images
  library(magick)

  # read scale bar image
  img_scalebar <- image_read("/home/sari.giering/git/plankton-dsg-challenge/data/collages/scalebar_100px.tif")[1]
  
  for(TYPE in levels(index$label3)) {
    
    # number of objects for TYPE
    n <- length(index$label3[index$label3 %in% TYPE])
    
    # load 4 largest and smallest images
    for(i in 1:4) {
      IMG_NAME <- paste("/scratch/data/images/", largest_5[largest_5$label3 %in% TYPE, ]$filename[i], sep="")
      img      <- image_read(IMG_NAME)
      assign(paste("img_L", i, sep="_"), img)

      IMG_NAME <- paste("/scratch/data/images/", smallest_5[smallest_5$label3 %in% TYPE, ]$filename[i], sep="")
      img      <- image_read(IMG_NAME)
      assign(paste("img_S", i, sep="_"), img)
    }
    
    # collate images
    col_L   <- image_append(c(img_L_1, img_L_2, img_L_3, img_L_4))
    col_S   <- image_append(c(img_S_1, img_S_2, img_S_3, img_S_4, img_scalebar))
    col_all <- image_append(c(col_L, col_S), stack = TRUE)
    
    # add annotation
    ANNOT <- paste(TYPE, " (n = ", n, ")", sep="")
    col_annotated <- image_annotate(col_all, ANNOT, size = 20, gravity = "northwest")
    
    # make filename
    FILENAME <- paste("/home/sari.giering/git/plankton-dsg-challenge/data/collages/collage_", TYPE,  "_n", n, ".png", sep = "")
    
    # ---- SAVE IMAGES ----
    #image_write(col_annotated, path = FILENAME, format = "png")
  }
   
```

# Class imbalances

I use this opportunity to illustrate the class imbalances. To add a bit more information to the plot, I also include information on the median file size, which is a proxy for object size.

The resulting plot nicely illustrates the variability in class size and class diversity of label3 images. Error bars show Q1 and Q3 (25% and 75% of data).

I commented out the table saving command here. Please enable if you want to save the table.


```{r show_class_imbalance}
  classes <- aggregate(index$filename, by = list("label1" = index$label1,
                                      "label2" = index$label2,
                                      "label3" = index$label3), length)

  classes$median_size <- aggregate(index$size, by = list("label1" = index$label1,
                                      "label2" = index$label2,
                                      "label3" = index$label3), median)$x

  classes$size_q1 <- aggregate(index$size, by = list("label1" = index$label1,
                                      "label2" = index$label2,
                                      "label3" = index$label3), quantile, probs = 0.25)$x
  
  classes$size_q3 <- aggregate(index$size, by = list("label1" = index$label1,
                                      "label2" = index$label2,
                                      "label3" = index$label3), quantile, probs = 0.75)$x
  
  library(ggplot2)
  ggplot(classes, aes(x = x, y = median_size/1000, color = label2)) +
      geom_point() +
      geom_errorbar(aes(ymin = size_q1/1000, ymax = size_q3/1000), width = 0.2) +
      scale_x_continuous(trans = 'log10') +
      theme_classic() + theme(legend.position=c(0.85, 0.75)) +
      ylab("Median file size (KB)") +
      xlab("Number of images in class") + ggtitle("Classes according to label3")
  
  knitr::kable(classes[rev(order(classes$x)),])
  
  # ---- SAVE TABLE ----
  #write.csv(classes[rev(order(classes$x)),], "out.csv", row.names = FALSE)
```